Gravitational-wave sensitivity curves

### 2014-09-24 V2/ReplyToReferees.txt

Dear Sir/Madam

Please find below a point by point reply to the comments of the two referees and a description of the changes we have made to the manuscript.

We have made changes to address all of the comments by the referees. These changes have removed a number of small typographical mistakes and improved the overall readability of the manuscript. All of the authors would like to thank the referees for their comments.

Yours faithfully,

Christopher Moore, Robert Cole and Christopher Berry.

Referee 1
It is very gratifying to hear that the referee enjoyed reading the manuscript. We would like to thank the referee for their time and their kind comments.

Referee 2
Major points:
1)	DISCUSS WITH ROB.

2)	We agree with this comment and think that it's well worth emphasising this point. We have added the following passage to the end of sec 5.4...

"It should be noted, however, that the sensitivity curves plotted in Appendix A would change substantially for stochastic backgrounds. This is because searches for stochastic backgrounds utilise the cross correlation of the outputs of multiple detectors for detection (Allen and Romano, 1999; Maggiore, 2000), instead of the individual outputs. Therefore the sensitivity of a network of ground based detectors is much greater than that of any individual detector. The shape of the sensitivity curve for a stochastic background would also differ from the curves plotted here, for a discussion of this point see Thrane and Romano (2013)."

Other comments (in order of appearance in the text):
1)	We thank the referee for pointing out this interesting source of confusion. We have checked this, and it seems that it was originally "evolved LISA", but the alternative "evolving LISA" has gradually become more popular and is now the prefered option. We have found numerous examples of both acronyms being used; for two conflicting examples on the same website see https://www.elisascience.org/news/top-news/gravitationaluniverseselectedasl3 and https://www.elisascience.org/articles/elisa-consortium/home . We have changed the main body of the text to read "evolving" added the following footnote...

"The “e” in eLISA originally stood for evolved, however, the preference is now for evolving."

2)	We agree with the referees comment and have amended the passage to read...

"Another disadvantage is that applying the definition 17 to a monochromatic source gives a formally undefined answer. The correct identification of characteristic strain for a monochromatic source is the amplitude of the wave times the square root of the number of periods observed."

3)	We have made this correction.

4)	We have made this correction.

5)	We have made this correction.

6)	We agree with the referee on this comment. The message we were trying to convey here is that current limits are comparable in order of magnitude. We have changed the relevant passage in sec 4.4.1 to read...

"The limits from the existing detectors are all of a similar magnitude; the PPTA limit is a factor of ∼ 2.5 lower than the other two. In Appendix A we use the EPTA limits, the figures show the published limit on A from equation 40 at a frequency of f0 = 1 yr−1 , based on an analysis of 5 pulsars over approximately 10 yr."

7)	THE EPTA CURVE DOESN'T ASSUME A DELTA_T. NEED TO CLARIFY ASSUMPTIONS FOR OTHER CURVES.

### 2014-09-24 V2/Detectors.tex
reply to referees in folder V2/

 \includegraphics[trim=0cm 0cm 0cm 0cm, width=0.6\textwidth]{Fig_Hellings_Downs}

@@

The PTAs currently in operation are the European Pulsar Timing Array \citep[EPTA\footnote{\url{http://www.epta.eu.org/}};][]{eptareview2013}, the Parkes Pulsar Timing Array \citep[PPTA\footnote{\url{http://www.atnf.csiro.au/research/pulsar/ppta/}};][]{parkesreview2013} based in Australia, and the North American Nanohertz Observatory for Gravitational waves \citep[NANOGrav\footnote{\url{http://nanograv.org/}};][]{nanogravreview2013}. There are published limits on the amplitude of the stochastic background from all three detectors: the most recent from EPTA is \citet{Haasteren}, from PPTA is \citet{Shannon2013} and from NANOGrav is \citet{2013ApJ...762...94D}. These are all comparable and we use the EPTA limits in \ref{app:a}, the figures show the published limit on $A$ from equation \ref{eq:power} at a frequency of $f_{0}=1~\mathrm{yr^{-1}}$, based on an analysis of $5$ pulsars over approximately $10~\mathrm{yr}$.

>>>>

 \includegraphics[trim=0cm 0cm 0cm 0cm, width=0.55\textwidth]{Fig_Hellings_Downs}

@@

The PTAs currently in operation are the European Pulsar Timing Array \citep[EPTA\footnote{\url{http://www.epta.eu.org/}};][]{eptareview2013}, the Parkes Pulsar Timing Array \citep[PPTA\footnote{\url{http://www.atnf.csiro.au/research/pulsar/ppta/}};][]{parkesreview2013} based in Australia, and the North American Nanohertz Observatory for Gravitational waves \citep[NANOGrav\footnote{\url{http://nanograv.org/}};][]{nanogravreview2013}. There are published limits on the amplitude of the stochastic background from all three detectors: the most recent from EPTA is \citet{Haasteren}, from PPTA is \citet{Shannon2013} and from NANOGrav is \citet{2013ApJ...762...94D}. The limits from the existing detectors are all of a similar magnitude; the PPTA limit is a factor of $\sim 2.5$ lower than the other two. In \ref{app:a} we use the EPTA limits, this was based on an analysis of $5$ pulsars over approximately $10~\mathrm{yr}$. The curve labelled ``EPTA'' is normalised to agree with the limit published in \cite{Haasteren} at a frequency of $f_{0}=1~\mathrm{yr^{-1}}$.

### 2014-09-24 V2/Introduction.tex
reply to referees in folder V2/

The next few years  promise to deliver the first direct detection of gravitational waves (GWs). This will most likely be achieved by the advanced versions of the Laser Interferometer Gravitational-wave Observatory \citep[LIGO;][]{2010CQGra..27h4006H} and Virgo \citep{Acernese2009} detectors operating in the frequency range $(10$--$10^{3})~\mathrm{Hz}$. By the end of the decade, it is expected that pulsar timing arrays \citep[PTAs;][]{1990ApJ361300F} will also detect very low frequency GWs around $10^{-8}~\mathrm{Hz}$. Further into the future, space-based detectors, such as the evolved Laser Interferometer Space Antenna \citep[eLISA;][]{TheGravitationalUniverse}, will probe GWs in the millihertz regime. These advances shall herald the beginning of multi-wavelength GW astronomy as a means of observing the Universe.

@@

An additional advantage of this convention is that the values on the strain axis for the detector curve $h_n(f)$ have a simple physical interpretation: they correspond to the root-mean-square noise in a bandwidth $f$. One downside to plotting characteristic strain is that the values on the strain axis $h_\mathrm{c}(f)$ do not directly relate to the amplitude of the waves from the source. Another disadvantage is that the characteristic strain for a monochromatic source is infinite.

@@

The PSD, as defined by (\ref{eq:psd}), has the nice property, demonstrated in (\ref{eq:meansquare2}), that integrated over all positive frequencies it gives the mean square amplitude of the signal in the detector. However, in one important regard it is less appealing than characteristic strain: the height of the source above the detector curve is no longer directly related to the SNR.

@@

The energy in GWs is described by the Isaacson stress--energy tensor \citep[section 35.15]{MTW}
\begin{equation}
T_{\mu\nu}=\frac{c^{4}}{32\pi G}\left<\partial_{\mu}\bar{h}_{\alpha\beta}\partial_{\nu}\bar{h}^{\alpha\beta}\right> \;,
\end{equation}
where the angle brackets denote averaging over several wavelengths or periods, and $\bar{h}_{\alpha\beta}$ is the transverse-traceless metric perturbation. The energy density $\rho c^{2}$ is given by the $T_{00}$ component of this tensor. Consequently \citep[cf.][]{Berry2013},
\begin{eqnarray}
\rho c^{2} &=& \frac{c^{2}}{16\pi G}\int_{-\infty}^{\infty}\mathrm{d}f\;\left(2\pi f\right)^{2}\tilde{h}(f)\tilde{h}^{*}(f) \\*
 &=& \int_{0}^{\infty}\mathrm{d}f\;\frac{\pi c^{2}}{4G}f^{2}S_{h}(f)\; , \label{eq:specNRGdensity}
\end{eqnarray} 
where the definition (\ref{eq:ShforSources}) have been used. The integrand in Equation (\ref{eq:specNRGdensity}) is defined as the spectral energy density, the energy per unit volume of space, per unit frequency \citep{HellingsDowns}
\begin{equation}\label{eq:spectralenergydensity}
S_{\mathrm{E}}(f)=\frac{\pi c^{2}}{4G} f^{2}S_{h}(f) \; ;
\end{equation}
a corresponding expression for the noise can be formulated by replacing $S_h(f)$ with $S_{n}(f)$.

>>>>

The next few years  promise to deliver the first direct detection of gravitational waves (GWs). This will most likely be achieved by the advanced versions of the Laser Interferometer Gravitational-wave Observatory \citep[LIGO;][]{2010CQGra..27h4006H} and Virgo \citep{Acernese2009} detectors operating in the frequency range $(10$--$10^{3})~\mathrm{Hz}$. By the end of the decade, it is expected that pulsar timing arrays \citep[PTAs;][]{1990ApJ361300F} will also detect very low frequency GWs around $10^{-8}~\mathrm{Hz}$. Further into the future, space-based detectors, such as the evolving Laser Interferometer Space Antenna \citep[eLISA;][]{TheGravitationalUniverse}
\footnote{The “e” in eLISA originally stood for evolved, however, the preference is now for evolving.}
These advances shall herald the beginning of multi-wavelength GW astronomy as a means of observing the Universe.

@@

An additional advantage of this convention is that the values on the strain axis for the detector curve $h_n(f)$ have a simple physical interpretation: they correspond to the root-mean-square noise in a bandwidth $f$. One downside to plotting characteristic strain is that the values on the strain axis $h_\mathrm{c}(f)$ do not directly relate to the amplitude of the waves from the source. 
Another disadvantage is that applying the definition \ref{eq:strain-hc} to a monochromatic source gives a formally undefined answer. The correct identification of characteristic strain for a monochromatic source is the amplitude of the wave times the square root of the number of periods observed.

%Another disadvantage is that the characteristic strain for a monochromatic source is infinite.

@@

The PSD, as defined by (\ref{eq:psd}), has the nice property, demonstrated in (\ref{eq:meansquare2}), that integrated over all positive frequencies it gives the mean square amplitude of the noise in the detector. However, in one important regard it is less appealing than characteristic strain: the height of the source above the detector curve is no longer directly related to the SNR.

@@

The energy in GWs is described by the Isaacson stress--energy tensor \citep[section 35.15]{MTW}
\begin{equation}
T_{\mu\nu}=\frac{c^{4}}{32\pi G}\left<\partial_{\mu}\bar{h}_{\alpha\beta}\partial_{\nu}\bar{h}^{\alpha\beta}\right> \;,
\end{equation}
where the angle brackets denote averaging over several wavelengths or periods, and $\bar{h}_{\alpha\beta}$ is the transverse-traceless metric perturbation. The energy density $\rho c^{2}$ is given by the $T_{00}$ component of this tensor. Consequently \citep[cf.][]{Berry2013},
\begin{eqnarray}
\rho c^{2} &=& \frac{c^{2}}{16\pi G}\int_{-\infty}^{\infty}\mathrm{d}f\;\left(2\pi f\right)^{2}\tilde{h}(f)\tilde{h}^{*}(f) \\*
 &=& \int_{0}^{\infty}\mathrm{d}f\;\frac{\pi c^{2}}{4G}f^{2}S_{h}(f)\; , \label{eq:specNRGdensity}
\end{eqnarray} 
where the definition (\ref{eq:ShforSources}) has been used. The integrand in Equation (\ref{eq:specNRGdensity}) is defined as the spectral energy density, the energy per unit volume of space, per unit frequency \citep{HellingsDowns}
\begin{equation}\label{eq:spectralenergydensity}
S_{\mathrm{E}}(f)=\frac{\pi c^{2}}{4G} f^{2}S_{h}(f) \; ;
\end{equation}
a corresponding expression for the noise can be formulated by replacing $S_h(f)$ with $S_{n}(f)$.

### 2014-09-24 V2/Sources.tex
reply to referees in folder V2/

>>>>

It should be noted, however, that the sensitivity curves plotted in \ref{app:a} would change substantially for stochastic backgrounds. This is because searches for stochastic backgrounds utilise the cross correlation of the outputs of multiple detectors for detection \citep{1999PhRvD..59j2001A,2000PhR...331..283M}, instead of the individual outputs. Therefore the sensitivity of a network of ground based detectors is much greater than that of any individual detector. The shape of the sensitivity curve for a stochastic background would also differ from the curves plotted here, for a discussion of this point see \cite{2013PhRvD..88l4032T}.

### 2014-09-24 V2/Typesofsource.tex
reply to referees in folder V2/

The form for $h_\mathrm{c}$ can be derived from the Fourier transform in the stationary-phase approximation. Consider a source signal with approximately constant (root-mean-square) amplitude $h_0$ and central frequency $f'$. In this case,
\begin{eqnarray}
h(t) &=& \sqrt{2} h_0 \cos\left[\phi(t)\right] \; , \\
\tilde{h}(f) &=& \frac{h_0}{\sqrt{2}} \int_{-\infty}^{\infty}\rmd t\; \exp \left\{2\pi \rmi \left[\frac{\phi(t)}{2\pi t}-f\right]t\right\} + \exp \left\{-2\pi \rmi \left[\frac{\phi(t)}{2\pi t}+f\right]t\right\} \; .
\label{eq:fourier-h}
\end{eqnarray}
Without loss of generality, we can assume an initial phase of zero, such that $\phi(0) = 0$. The largest contribution to the integral comes from where the argument of the exponentials is approximately zero. For the first term, this occurs when $f = f'$, then the term in brackets is
\begin{equation}
\left[\frac{\phi(t)}{2\pi t}-f\right]_{f\,=\,f'} = f' + \dot{f'}t + \mathcal{O}\left(t^2\right) - f' = \dot{f'}t + \mathcal{O}\left(t^2\right).
\end{equation}
The higher-order terms cause the expoential to oscillate rapidly such that these terms integrate to zero and may be neglected. Performing a similar expansion about $f = -f'$ for the second term in (\ref{eq:fourier-h}) and then evaluating the Gaussian integrals gives
\begin{eqnarray} \label{eq:FTofEMRI}
\tilde{h}(f) &\simeq& \frac{h_0}{\sqrt{2}} \int_{-\infty}^{\infty}\rmd t\; \exp\left(2\pi\rmi \dot{f'} t^{2}\right) + \exp\left(-2\pi\rmi \dot{f'} t^{2}\right) \nonumber \\
&\simeq& \frac{h_0}{\sqrt{2\dot{f'}}}.
\end{eqnarray}
From (\ref{eq:strain-hc}) and (\ref{eq:FTofEMRI}), the characteristic strain for inspiralling sources is given by \citep{FinnThorne}
\begin{equation}\label{eq:insphc}
h_\mathrm{c}(f) = \sqrt{\frac{2f^{2}}{\dot{f}}}h_0 \;.
\end{equation}
Equation (\ref{eq:hc}) should be considered as the definition of characteristic strain and (\ref{eq:insphc}) a consequence of it for inspirals. Equation (\ref{eq:insphc}) is the relation between $h_\mathrm{c}(f)$ and the instantaneous root-mean-square amplitude $h_0$ for an inspiralling source; for other types of source a new relation satisfying (\ref{eq:hc}) has to be found.

>>>>

The form for $h_\mathrm{c}$ can be derived from the Fourier transform in the stationary-phase approximation. Consider a source signal with approximately constant (root-mean-square) amplitude $h_0$ and central frequency $f'$. In this case,
\begin{eqnarray}
h(t) &=& \sqrt{2} h_0 \cos\left[\phi(t)\right] \; , \\
\tilde{h}(f) &=& \frac{h_0}{\sqrt{2}} \int_{-\infty}^{\infty}\rmd t\; \exp \left\{2\pi \rmi \left[\frac{\phi(t)}{2\pi t}-f\right]t\right\} + \exp \left\{-2\pi \rmi \left[\frac{\phi(t)}{2\pi t}+f\right]t\right\} \; .
\label{eq:fourier-h}
\end{eqnarray}
Without loss of generality, we can assume an initial phase of zero, such that $\phi(0) = 0$. The largest contribution to the integral comes from where the argument of the exponentials is approximately zero. For the first term, this occurs when $f = f'$, then the term in brackets is
\begin{equation}
\left[\frac{\phi(t)}{2\pi t}-f\right]_{f\,=\,f'} = f' + \dot{f'}t + \mathcal{O}\left(t^2\right) - f' = \dot{f'}t + \mathcal{O}\left(t^2\right).
\end{equation}
The higher-order terms cause the expoential to oscillate rapidly such that these terms integrate to zero and may be neglected. Performing a similar expansion about $f = -f'$ for the second term in (\ref{eq:fourier-h}) and then evaluating the Gaussian integrals gives
\begin{eqnarray} \label{eq:FTofEMRI}
\tilde{h}(f) &\simeq& \frac{h_0}{\sqrt{2}} \int_{-\infty}^{\infty}\rmd t\; \exp\left(2\pi\rmi \dot{f'} t^{2}\right) + \exp\left(-2\pi\rmi \dot{f'} t^{2}\right) \nonumber \\
&\simeq& \frac{h_0}{\sqrt{2\dot{f'}}}.
\end{eqnarray}
From (\ref{eq:strain-hc}) and (\ref{eq:FTofEMRI}), the characteristic strain for inspiralling sources is given by \citep{FinnThorne}
\begin{equation}\label{eq:insphc}
h_\mathrm{c}(f) = \sqrt{\frac{2f^{2}}{\dot{f}}}h_0 \;.
\end{equation}
Equation (\ref{eq:strain-hc}) should be considered as the definition of characteristic strain and (\ref{eq:insphc}) a consequence of it for inspirals. Equation (\ref{eq:insphc}) is the relation between $h_\mathrm{c}(f)$ and the instantaneous root-mean-square amplitude $h_0$ for an inspiralling source; for other types of source a new relation satisfying (\ref{eq:hc}) has to be found.
